---
title: "miRNA-Analysis"
author: "Evan"
date: '2022-10-24'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading in Data

#### Reading in HLHS-CPC EV miRNA Data from Davis Lab
```{r}
library(limma)
library(Glimma)
library(edgeR)
library(readxl)
library(tidyverse)
library(readxl)
library(readr)
library(dplyr)
#reading in csv data and converting to matrix
DavisEV<-read.csv("miRNA_counts_EV.csv", header = TRUE)%>%
  column_to_rownames(var="X")

#selecting proper samples
DavisEV<-dplyr::select(as.data.frame(DavisEV),c("E1101","E1102","E1103","E1104","E1105","E1106","E1107"))

#removing all columns with less than 50% zeroes
DavisEV<-DavisEV[apply(DavisEV == 0, 1, sum) <= ncol(DavisEV)/2, ]

DavisEV<-na.omit(DavisEV)
DavisDGE<-DGEList(counts=DavisEV)
#grouping annotations
group <- as.factor(c("HLHS_CPC-EV"))
DavisDGE$samples$group <- group


```

#### Reading in Endothelial Progenitor Cells EV mi-RNA Data
From this: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE124692
Note- we used miRNA count data at the link above

```{r}
EPCmiRNAData<- read_excel("miRNA_EPC.xlsx")
EPCmiRNAData<- EPCmiRNAData[,colnames(EPCmiRNAData)!="miRNApreName"]
EPCmiRNAData<- EPCmiRNAData[,colnames(EPCmiRNAData)!="mirSequence"]
EPCData<- as.data.frame(EPCmiRNAData)
rownames(EPCData)<- EPCData[,1]
EPCData= EPCData[,-1]


EPCData<-EPCData[apply(EPCData == 0, 1, sum) <= ncol(EPCData)/2, ]

EPCDGE<-DGEList(counts=EPCData)

#grouping annotations
group <- as.factor(c("miRNA-EPC-EV"))
EPCDGE$samples$group <- group


```


#### Reading in MSC-EV miRNA Data
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE159814 

```{r}
MSCData<- read_excel("GSE159814_miRNA_readscount.xlsx")
MSCData<-as.data.frame(MSCData)
rownames(MSCData)<- MSCData[,1]
MSCData= MSCData[,-1]

MSCDGE<-DGEList(counts=MSCData)
#grouping annotations
group <- as.factor(c("miRNA-MSC-EV"))
MSCDGE$samples$group <- group


```


#### Merging All Datasets

```{r}

mergeEPC<-merge(DavisDGE$counts,EPCDGE$counts,by=0)

mergeMSC<-merge(DavisDGE$counts,MSCDGE$counts,by=0)

#mergedData<-merge(merge1,MSCDGE$counts,by=0)

#view(merge1)
#view(MSCDGE$counts)
#view(mergedData)


```
## Analyzing mergeEPC and generating top table

```{r}
cleanMergedEPCDF <- mergeEPC[,-1]
rownames(cleanMergedEPCDF) <- mergeEPC[,1]

mergedEPCMatrix<-data.matrix(cleanMergedEPCDF)

mergedEPCDGE<-DGEList(counts=mergedEPCMatrix)

#adding gene annotations
group <- as.factor(rep(c("HLHS_CPC","EPC"), c(7,4)))
mergedEPCDGE$samples$group<-group

#Analysis
cpm<-cpm(mergedEPCDGE)
lcpm<-cpm(mergedEPCDGE,log=TRUE)
#L and M values
L <- mean(mergedEPCDGE$samples$lib.size) * 1e-6
M <- median(mergedEPCDGE$samples$lib.size) * 1e-6
c(L, M)

#removing lowly expressed genes
samplenames<-colnames(mergedEPCDGE)
keep.exprs <- filterByExpr(mergedEPCDGE, group=group)
mergedEPCDGE <- mergedEPCDGE[keep.exprs,, keep.lib.sizes=FALSE]
dim(mergedEPCDGE)


```

```{r}
#QC1
lcpm.cutoff <- log2(10/M + 2/L)
library(RColorBrewer)
nsamples <- ncol(mergedEPCDGE)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
lcpm <- cpm(mergedEPCDGE, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")


```
```{r}
mergedEPCDGE <- calcNormFactors(mergedEPCDGE, method = "TMM")
mergedEPCDGE$samples$norm.factors
lcpm <- cpm(mergedEPCDGE, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="Boxplot for Data",ylab="Log-cpm")

```
```{r}
#QC3
lcpm <- cpm(mergedEPCDGE, log=TRUE)
par(mfrow=c(1,2))
col.group <- group
levels(col.group) <-  brewer.pal(nlevels(col.group), "Set1")
col.group <- as.character(col.group)

plotMDS(lcpm, labels=group, col=col.group)
title(main="A. Sample groups")

```
## Differential Expression Analysis for EPC vs HLHS CPC
```{r}
design <- model.matrix(~0+group)
colnames(design) <- gsub("group", "", colnames(design))
design

v <- voom(mergedEPCDGE, design, plot=TRUE)
v

contr.matrix <- makeContrasts(
   HLHS_CPC=HLHS_CPC, 
   EPC=EPC,
   levels = colnames(design))
contr.matrix

vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=contr.matrix)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")

summary(decideTests(efit))
#the below method is stricter in terms of decisions
tfit <- treat(vfit, lfc=1)
dt <- decideTests(tfit)
summary(dt)

```
```{r}
export <- topTreat(tfit,coef=1,n=Inf)
export<-filter(as.data.frame(export),export$adj.P.Val<0.01)
export<-cbind(rownames(export),export)
view(export)
```
```{r}
library("writexl")
write_xlsx(export,"HLHS-CPC_vs_EPC.xlsx")
```

#### Analyzing mergeMSC and generating top table

```{r}
cleanMergedMSCDF <- mergeMSC[,-1]
rownames(cleanMergedMSCDF) <- mergeMSC[,1]

mergedMSCMatrix<-data.matrix(cleanMergedMSCDF)

mergedMSCDGE<-DGEList(counts=mergedMSCMatrix)

#adding gene annotations
group <- as.factor(rep(c("HLHS_CPC","MSC"), c(7,3)))
mergedMSCDGE$samples$group<-group

#Analysis
cpm<-cpm(mergedMSCDGE)
lcpm<-cpm(mergedMSCDGE,log=TRUE)
#L and M values
L <- mean(mergedMSCDGE$samples$lib.size) * 1e-6
M <- median(mergedMSCDGE$samples$lib.size) * 1e-6
c(L, M)

#removing lowly expressed genes
samplenames<-colnames(mergedMSCDGE)
keep.exprs <- filterByExpr(mergedMSCDGE, group=group)
mergedMSCDGE <- mergedMSCDGE[keep.exprs,, keep.lib.sizes=FALSE]
dim(mergedMSCDGE)


```

```{r}
#QC1
lcpm.cutoff <- log2(10/M + 2/L)
library(RColorBrewer)
nsamples <- ncol(mergedMSCDGE)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
lcpm <- cpm(mergedMSCDGE, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")


```

```{r}
mergedMSCDGE <- calcNormFactors(mergedMSCDGE, method = "TMM")
mergedMSCDGE$samples$norm.factors
lcpm <- cpm(mergedMSCDGE, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="Boxplot for Data",ylab="Log-cpm")

```
```{r}
#QC3
lcpm <- cpm(mergedMSCDGE, log=TRUE)
par(mfrow=c(1,2))
col.group <- group
levels(col.group) <-  brewer.pal(nlevels(col.group), "Set1")
col.group <- as.character(col.group)

plotMDS(lcpm, labels=group, col=col.group)
title(main="A. Sample groups")

```
## Differential Expression Analysis for MSC vs HLHS CPC
```{r}
design <- model.matrix(~0+group)
colnames(design) <- gsub("group", "", colnames(design))
design

v <- voom(mergedMSCDGE, design, plot=TRUE)
v

contr.matrix <- makeContrasts(
   HLHS_CPC=HLHS_CPC, 
   MSC=MSC,
   levels = colnames(design))
contr.matrix

vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=contr.matrix)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")

summary(decideTests(efit))
#the below method is stricter in terms of decisions
tfit <- treat(vfit, lfc=1)
dt <- decideTests(tfit)
summary(dt)

```
```{r}
export <- topTreat(tfit,coef=1,n=Inf)
export<-filter(as.data.frame(export),export$adj.P.Val<0.01)
export<-cbind(rownames(export),export)
view(export)
```
```{r}
library("writexl")
write_xlsx(export,"HLHS-CPC_vs_MSC.xlsx")
```


#### Ignore Below
```{r}
#put all data frames into list
df_list <- list(DavisDGE$counts, EPCDGE$counts, MSCDGE$counts)

#merge all data frames in list
#Reduce(function(x, y) merge(x, y, all=TRUE), df_list)

#merge all data frames in list
df_list %>% reduce(full_join, by='variable_name')

```